---
title: "Ski Resort Data Analysis - Final Paper"
author: "Arnav, Sam, Quinn, and Santiago"
date: today
format: pdf
editor: source
---


```{r, message=FALSE}
#| label: Installing-packages
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(dplyr)
library(knitr)
library(reshape2)
set.seed(123)
```

# Reading and Cleaning Data
```{r}
#| label: reading-data
resort_df <- read_csv("resorts.csv")
snow_df <- read_csv("snow.csv")
metadata <- read_csv("data_dictionary.csv")

head(resort_df)
head(snow_df)
head(metadata)

```


Dataset was found on https://www.mavenanalytics.io/data-playground?order=date_added%2Cdesc&search=ski%20resort. The csvs have been generated through webscraping from https://ski-resort-stats.com/ and the snow data come from https://nsidc.org/data/mod10cm/versions/61.

```{r}
#| label: cleaning-data
clean_non_alphanumeric <- function(data_col) {
  # Convert non-ASCII characters to ASCII
  data_col <- iconv(data_col, from = "UTF-8", to = "ASCII", sub = "")
  # Remove non-alphanumeric characters except hyphens
  data_col <- gsub("[^a-zA-Z0-9 \\-]", "", data_col)
  # Remove extra spaces
  data_col <- gsub("\\s+", " ", data_col)
  # Trim leading and trailing spaces
  trimws(data_col)
}

# Create a new dataframe with only the cleaned Resort column
cleaned_resort_df <- data.frame(CleanedResort = clean_non_alphanumeric(resort_df$Resort))

# Add the original Resort names for comparison
cleaned_resort_df$OriginalResort <- resort_df$Resort

# Write the cleaned data to a new CSV file
write.csv(cleaned_resort_df, "cleaned_resort_names.csv", row.names = FALSE)
```

```{r}
#| label: cleaning-column-names
# Rename ID column to Resort_ID
resort_df <- resort_df |>
  rename(Resort_ID = ID)

resort_df <- resort_df |>
  mutate(clean_resort_name = clean_non_alphanumeric(Resort))
```


```{r}
#| label: fixing-missing-incorrect-values

#Removing NAs and weird values
resort_df <- resort_df |> filter(Price > 0)

# Some ski resorts have "Unknown" for their ski season. For these we set a sensible value if they are in the northern or southern hemisphere

resort_df <- resort_df %>%
  mutate(Hemisphere = ifelse(Latitude > 0, "Northern", "Southern"))

# Function to process Season column
process_season <- function(season) {
  if (is.na(season)) return(NULL)
  
  # Handle "Year-round"
  if (tolower(season) == "year-round") {
    return(data.frame(Start = 1, End = 12))
  } else if (tolower(season=="unknown")) {
    return(data.frame(Start = NA, End = NA))
  }
  
  # Split multiple intervals
  intervals <- strsplit(season, ",\\s*")[[1]]
  
  # Parse each interval
  parsed_intervals <- lapply(intervals, function(interval) {
    months <- str_extract_all(interval, "[A-Za-z]+")[[1]]
    if (length(months) == 1) {
      # Single-month season
      start <- match(months, month.name)
      return(data.frame(Start = start, End = start))
    } else if (length(months) == 2) {
      # Range of months
      start <- match(months[1], month.name)
      end <- match(months[2], month.name)
      return(data.frame(Start = start, End = end))
    }
    return(NULL)
  })
  
  # Combine parsed intervals
  bind_rows(parsed_intervals)
}

# Process all seasons and expand intervals
season_data <- resort_df %>%
  mutate(
    Season_Data = map(Season, process_season)
  ) %>%
  unnest(Season_Data)

# Calculate average season start and end by hemisphere
average_season_hemisphere <- season_data %>%
  group_by(Hemisphere) %>%
  summarise(
    Avg_Season_Start = mean(Start, na.rm = TRUE),
    Avg_Season_End = mean(End, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Avg_Season = paste(month.name[round(Avg_Season_Start)], "-", month.name[round(Avg_Season_End)])
  )


# Calculate average season start and end by continent
average_season_continent <- season_data %>%
  group_by(Continent) %>%
  summarise(
    Avg_Season_Start = mean(Start, na.rm = TRUE),
    Avg_Season_End = mean(End, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Avg_Season = paste(month.name[round(Avg_Season_Start)], "-", month.name[round(Avg_Season_End)])
  )

# Print results
print("Average season by hemisphere:")
print(average_season_hemisphere)

print("Average season by continent:")
print(average_season_continent)

# Extract the average season for each hemisphere
northern_avg_season <- average_season_hemisphere %>%
  filter(Hemisphere == "Northern") %>%
  pull(Avg_Season)

southern_avg_season <- average_season_hemisphere %>%
  filter(Hemisphere == "Southern") %>%
  pull(Avg_Season)

# Assign default seasons to "Unknown" based on hemisphere averages
resort_df <- resort_df %>%
  mutate(
    Season = case_when(
      Season == "Unknown" & Hemisphere == "Northern" ~ northern_avg_season,
      Season == "Unknown" & Hemisphere == "Southern" ~ southern_avg_season,
      Season == "Year-round" ~ "January - December",
      TRUE ~ Season  # Keep other seasons as is
    )
  )
#Now resort_df should have no missing or incorrect values, and seasons are now consistent

```



```{r}
#| label: merging-IDs
# Function to assign resort IDs based on latitude and longitude with a specified tolerance
assign_resort_ids <- function(snow_data, resort_data, tolerance = 0.1) {
  # Initialize Resort_ID as a character column with empty strings
  snow_data <- snow_data %>%
    mutate(Resort_ID = "")
  
  for (i in seq_len(nrow(resort_data))) {
    # Get the latitude and longitude of the current resort
    resort_lat <- resort_data$Latitude[i]
    resort_lon <- resort_data$Longitude[i]
    resort_id <- resort_data$Resort_ID[i]
    
    # Filter snow data within the specified tolerance
    snow_data <- snow_data %>%
      mutate(
        Resort_ID = ifelse(
          Latitude >= (resort_lat - tolerance) & Latitude <= (resort_lat + tolerance) &
          Longitude >= (resort_lon - tolerance) & Longitude <= (resort_lon + tolerance),
          ifelse(
            Resort_ID == "",
            as.character(resort_id),  # Assign the first ID
            paste0(Resort_ID, ", ", resort_id)  # Append additional IDs
          ),
          Resort_ID
        )
      )
  }
  
  # Replace empty strings with NA for unmatched rows
  snow_data <- snow_data %>%
    mutate(Resort_ID = ifelse(Resort_ID == "", NA, Resort_ID))
  
  return(snow_data)
}

# Specify file paths for saving/loading
snow_df2_file <- "snow_df2.csv"
avg_snow_file <- "avg_snow_by_resort_and_month.csv"

# Check if `snow_df2` already exists
if (!file.exists(snow_df2_file)) {
  # Create `snow_df2` with Resort_ID assignment
  snow_df2 <- snow_df %>%
    mutate(Resort_ID = NA)
  
  # Assign Resort_IDs to snow_df
  tolerance_value <- 0.5  # Set your desired tolerance
  snow_df2 <- assign_resort_ids(snow_df2, resort_df, tolerance = tolerance_value)
  
  # Save `snow_df2` to a CSV file
  write_csv(snow_df2, snow_df2_file)
} else {
  # Load `snow_df2` from CSV
  snow_df2 <- read_csv(snow_df2_file)
}

#removing snow readings that aren't near a resort to reduce computation
snow_df_with_ID_no_NA <- snow_df2 |> filter(!is.na(Resort_ID))

# Ensure Resort_ID is a list of IDs split by commas
snow_df_with_ID_no_NA <- snow_df_with_ID_no_NA %>%
  mutate(
    Resort_ID = strsplit(as.character(Resort_ID), ",\\s*")  # Split IDs into lists
  )

# Expand snow_df_with_ID_no_NA to have one row per Resort_ID
expanded_snow_df <- snow_df_with_ID_no_NA %>%
  unnest(Resort_ID) %>%
  mutate(Resort_ID = as.numeric(Resort_ID))  # Convert Resort_ID to numeric for proper grouping

# expanded snow df contains all snow coverage readings within 55km of each resort over all 12 months
# we will average this so that we have an average snow coverage reading per month for each resort
# note some resorts may not have an average reading for some months


# Create `avg_snow_by_resort_and_month` dataframe
if (!file.exists(avg_snow_file)) {
  
  # Group by Resort_ID and Month, and compute the average snow
  avg_snow_by_resort_and_month <- expanded_snow_df %>%
    group_by(Resort_ID, Month) %>%
    summarise(
      avg_snow = mean(Snow, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(Resort_ID, Month)

  
  # Save `avg_snow_by_resort_and_month` to a CSV file
  write_csv(avg_snow_by_resort_and_month, avg_snow_file)
} else {
  avg_snow_by_resort_and_month <- read_csv(avg_snow_file)
}

# Ensure Resort_ID is numeric in avg_snow_by_resort_and_month
all_resort_ids <- unique(resort_df$Resort_ID)

# Find missing Resort_IDs
missing_resort_ids <- setdiff(all_resort_ids, unique(avg_snow_by_resort_and_month$Resort_ID))

# Print the missing IDs
print(missing_resort_ids) # This is the resort that has no snow data

```
Using a radius of 0.5 degrees in latitude and longitude (equivalent to ~55km at the equator), we determined the average snow for each resort location for each month. Interestingly some resorts (especially in the europe region) are clustered very close together - within this 55km radius. As such, for the weather stations that lie in an overlapping region between 2 nearby resorts we have an average snow reading that applied to both resorts. For each resort we take all weather stations within our tolerance (±0.5 deg latitude and longitude) and average their readings for each month.

Some resorts don't have snow coverage readings for all months. This is presumably due to outages for weather cells located in those areas.



## Data Analysis Plots

### Resort location on world map and snow readings


```{r}
#| label: plotting-resort-and-snow-on-map
#plot the latitude and longitude of resorts as well as the latitude and longitude in the snow dataset with a different color
# Extract unique latitude and longitude from the snow dataset
unique_snow_df <- snow_df %>%
  distinct(Latitude, Longitude)

# Plot the latitude and longitude of resorts and snow observation locations over a world map
ggplot() +
  # Add the world map as a base layer
  borders("world", colour = "gray95", fill = "gray95") +
  coord_fixed(1.3) + # Set aspect ratio for better map projection
  # Plot snow observation locations
  geom_point(data = unique_snow_df, aes(x = Longitude, y = Latitude, color = "Snow Observation Locations"), 
             alpha = 0.7, size = 2, shape = 16) +
  # Plot resort locations
  geom_point(data = resort_df, aes(x = Longitude, y = Latitude, color = "Resort Locations"), 
             alpha = 0.8, size = 2, shape = 17) +
  # Customize colors for points
  scale_color_manual(
    values = c("Resort Locations" = "blue", "Snow Observation Locations" = "lightblue"),
    name = "Dataset"
  ) +
  # Add labels and styling
  labs(
    title = "Global Map of Resorts and Snow Observation Locations",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```
It is clear that there are many more snow weather observation stations across the whole world than just at the resort locations. When merging the 2 dataframes we only need to consider the snow observations for gps locations where there is a resort, so we will assign these observations with an ID that corresponds to the resort that is being measured. For snow observations at a longitude or latitude that don't contain a resort will have an ID value of NaN.


### Average Ski Season Start and End by Continent
```{r, fig.height=5, fig.width=4}
# Create the boxplots for start and end
ggplot(season_data, aes(x = Continent, y = Start, fill = Hemisphere)) +
  geom_jitter(aes(color = Hemisphere), width = 0.3, alpha = 0.5, size = 0.5) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA, coef=2) +
  labs(
    title = "Ski Season Start by Continent",
    x = "Continent",
    y = "Season Start (Month)",
    fill = "Hemisphere",
    color = "Hemisphere",
    caption = "Boxplots show the distribution of ski season start months across continents."
  ) +
  scale_y_continuous(
    breaks = 1:12,
    labels = month.name,
    limits = c(0, 13)
  ) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2") +
  coord_cartesian(expand = FALSE) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.caption = element_text(size=4),
    legend.position = "top"
  )

ggplot(season_data, aes(x = Continent, y = End, fill = Hemisphere)) +
  geom_jitter(aes(color = Hemisphere), width = 0.3, alpha = 0.5, size = 0.5) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA, coef=4) +
  labs(
    title = "Ski Season End by Continent",
    x = "Continent",
    y = "Season End (Month)",
    fill = "Hemisphere",
    color = "Hemisphere",
    caption = "Boxplots show the distribution of ski season end months across continents."
  ) +
  scale_y_continuous(
    breaks = 1:12,
    labels = month.name,
    limits = c(0, 13)
  ) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2") +
  coord_cartesian(expand = FALSE) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.caption = element_text(size=4),
    legend.position = "top",
  )

```
### Correlation Heatmap
```{r}

# Filter numeric columns excluding Latitude, Longitude, and Resort_ID
numeric_columns <- resort_df %>%
  select_if(is.numeric) %>%
  select(-Latitude, -Longitude, -Resort_ID)

# Calculate the correlation matrix
correlation_matrix <- cor(numeric_columns, use = "pairwise.complete.obs")

# Melt the correlation matrix for ggplot2
melted_correlation <- melt(correlation_matrix, na.rm = TRUE)

# Plot the heatmap for the lower triangular matrix
ggplot(melted_correlation, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +  # Heatmap tiles
  geom_text(aes(label = round(value, 2)), color = "black", size = 3) +  # Add correlation values
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white",
    midpoint = 0, limit = c(-1, 1), space = "Lab",
    name = "Correlation"
  ) +
  theme_minimal() +
  labs(
    title = "Correlation Heatmap of Ski Resort Data",
    x = "",
    y = ""
  ) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

```

### Snow coverage by month for each continent
```{r}
# Join avg_snow_by_resort_and_month with resort_df, selecting only Continent and ID columns
avg_snow_with_continent <- avg_snow_by_resort_and_month %>%
  left_join(resort_df %>% select(Resort_ID, Continent), 
            by = c("Resort_ID" = "Resort_ID"))

# Summarize average snow coverage per month for each continent
avg_snow_by_continent <- avg_snow_with_continent %>%
  group_by(Continent, Month) %>%
  summarise(
    avg_snow_coverage = mean(avg_snow, na.rm = TRUE),  # Calculate average snow coverage
    .groups = "drop"
  )

# Convert Month to a proper date format for better plotting
avg_snow_by_continent <- avg_snow_by_continent %>%
  mutate(Month = as.Date(Month, format = "%Y-%m-%d"))

# Plot the line plot showing average snow coverage per month by continent
ggplot(avg_snow_by_continent, aes(x = Month, y = avg_snow_coverage, color = Continent, group = Continent)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Average Snow Coverage Per Month by Continent",
    x = "Month",
    y = "Average Snow Coverage (cm)",
    color = "Continent"
  ) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +  # Show month names
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

### Linear Regression

```{r}

# One-hot encode the Continent column
continent_one_hot <- model.matrix(~ Continent + 0, data = resort_df) %>%
  as.data.frame()

# Rename columns for clarity
colnames(continent_one_hot) <- gsub("Continent", "", colnames(continent_one_hot))

# Add the one-hot encoded columns back to the dataframe
resort_df_with_encoded_continents <- bind_cols(resort_df, continent_one_hot)

# Split the data into training and testing sets (80% training, 20% testing)
data_split <- initial_split(resort_df_with_encoded_continents, prop = 0.8)
training_data <- training(data_split)
testing_data <- testing(data_split)

# Define a linear regression model specification
lm_spec <- linear_reg() %>%
  set_engine("lm")

# Fit the model using the training data
lm_fit <- lm_spec %>%
  fit(
    Price ~ `Highest point` + `Difficult slopes` + `Lowest point` + `Beginner slopes` + `Intermediate slopes` +
      `Total lifts` + `Longest run` + `Snow cannons` + 
      `Europe` + `Asia` + `North America` + `Oceania`,
    data = training_data
  )

# Summarize the model
summary(lm_fit$fit)

# Predict on the training and testing sets
training_data <- training_data %>%
  mutate(
    Predicted_Price_LM = predict(lm_fit, new_data = training_data)$.pred
  )
testing_data <- testing_data %>%
  mutate(
    Predicted_Price_LM = predict(lm_fit, new_data = testing_data)$.pred
  )

# Calculate average price by continent using the training data
continent_avg_price <- training_data %>%
  group_by(Continent) %>%
  summarise(Avg_Price = mean(Price, na.rm = TRUE), .groups = "drop")

# Add continent average price predictions to both datasets
training_data <- training_data %>%
  left_join(continent_avg_price, by = "Continent") %>%
  mutate(
    Predicted_Price_Continent = Avg_Price
  )
testing_data <- testing_data %>%
  left_join(continent_avg_price, by = "Continent") %>%
  mutate(
    Predicted_Price_Continent = Avg_Price
  )

# Calculate RMSE for both models on testing data
rmse_lm <- rmse(testing_data, truth = Price, estimate = Predicted_Price_LM)
rmse_continent <- rmse(testing_data, truth = Price, estimate = Predicted_Price_Continent)

# Print the RMSE values
cat("RMSE for Linear Regression Model (Testing Data):", rmse_lm$.estimate, "\n")
cat("RMSE for Continent Average Model (Testing Data):", rmse_continent$.estimate, "\n")

```




```{r}
# Combine training and testing data for overall residual analysis
resort_df_with_predictions <- bind_rows(
  training_data %>% mutate(Data_Set = "Training"),
  testing_data %>% mutate(Data_Set = "Testing")
)

# Add residuals for both models
resort_df_with_predictions <- resort_df_with_predictions %>%
  mutate(
    Residual_LM = Price - Predicted_Price_LM,
    Residual_Continent = Price - Predicted_Price_Continent
  )

# Residual plot for Linear Regression Model
ggplot(resort_df_with_predictions, aes(x = Predicted_Price_LM, y = Residual_LM, color = Data_Set)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot for Linear Regression Model (All Data)",
    x = "Predicted Price (Linear Regression)",
    y = "Residuals",
    color = "Data Set"
  ) +
  theme_minimal()

# Residual plot for Continent Average Model
ggplot(resort_df_with_predictions, aes(x = Predicted_Price_Continent, y = Residual_Continent, color = Data_Set)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot for Continent Average Model (All Data)",
    x = "Predicted Price (Continent Average)",
    y = "Residuals",
    color = "Data Set"
  ) +
  theme_minimal()

# Calculate RMSE for both models
rmse_lm_all <- sqrt(mean(resort_df_with_predictions$Residual_LM^2, na.rm = TRUE))
rmse_continent_all <- sqrt(mean(resort_df_with_predictions$Residual_Continent^2, na.rm = TRUE))

# Display a comparison table
comparison_df <- data.frame(
  Model = c("Linear Regression", "Continent Average"),
  RMSE = c(rmse_lm_all, rmse_continent_all)
)

# Print the table with knitr
kable(
  comparison_df,
  caption = "Comparison of RMSE Between Linear Regression and Continent Average Models (All Data)"
)
```

```{r}
# Combine actual and predicted values for both models into long format
long_predictions <- resort_df_with_predictions %>%
  select(Price, Predicted_Price_LM, Predicted_Price_Continent) %>%
  pivot_longer(
    cols = c(Predicted_Price_LM, Predicted_Price_Continent),
    names_to = "Model",
    values_to = "Predicted_Price"
  ) %>%
  mutate(Model = recode(Model, 
                        Predicted_Price_LM = "Linear Regression",
                        Predicted_Price_Continent = "Continent Average"))

# Scatter plot: Actual vs Predicted
ggplot(long_predictions, aes(x = Price, y = Predicted_Price, color = Model)) +
  geom_point(alpha = 0.8,size=1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  labs(
    title = "Actual vs Predicted Prices by Model",
    x = "Actual Price",
    y = "Predicted Price",
    color = "Model"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "top"
  )
```

```{r}
# Combine residuals for both models into long format
long_residuals <- resort_df_with_predictions %>%
  select(Residual_LM, Residual_Continent) %>%
  pivot_longer(
    cols = c(Residual_LM, Residual_Continent),
    names_to = "Model",
    values_to = "Residual"
  ) %>%
  mutate(Model = recode(Model, 
                        Residual_LM = "Linear Regression",
                        Residual_Continent = "Continent Average"))

# Density plot for residuals
ggplot(long_residuals, aes(x = Residual, fill = Model)) +
  geom_density(alpha = 0.6) +
  labs(
    title = "Distribution of Residuals by Model",
    x = "Residual",
    y = "Density",
    fill = "Model"
  ) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "top"
  )
```


