---
title: "Ski Resort Data Analysis - Santiago"
author: "Santiago Gomez Alarcon"
format: pdf
editor: source
date: today
---

```{r, message=FALSE}
#| label: Installing-packages
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(dplyr)
library(knitr)
library(geosphere)  
library(ggrepel)
library(broom)
```

# Reading and Cleaning Data
```{r}
#| label: reading-data
resort_df <- read_csv("/Users/santi/ds1_2930/HD2930_Final_Project/resorts.csv")
snow_df <- read_csv("/Users/santi/ds1_2930/HD2930_Final_Project/snow.csv")
metadata <- read_csv("/Users/santi/ds1_2930/HD2930_Final_Project/data_dictionary.csv")

head(snow_df)
head(metadata)
head(resort_df)
```

Dataset was found on https://www.mavenanalytics.io/data-playground?order=date_added%2Cdesc&search=ski%20resort. The csvs have been generated through webscraping from https://ski-resort-stats.com/ and the snow data come from https://nsidc.org/data/mod10cm/versions/61.

```{r}
# Define the cleaning function
clean_resort_alphanumeric <- function(data_col) {
  # Convert non-ASCII characters to ASCII
  data_col <- iconv(data_col, from = "UTF-8", to = "ASCII", sub = "")
  # Remove non-alphanumeric characters except hyphens
  data_col <- gsub("[^a-zA-Z0-9 \\-]", "", data_col)
  # Remove extra spaces
  data_col <- gsub("\\s+", " ", data_col)
  # Trim leading and trailing spaces
  data_col <- trimws(data_col)
  return(data_col)
}

# Clean the Resort column in the original dataframe
resort_df$Resort <- clean_resort_alphanumeric(resort_df$Resort)

# Create a new dataframe with the cleaned Resort column replacing the original
cleaned_resort_df <- resort_df

# Save the new dataframe with cleaned resort names to a CSV file
write.csv(cleaned_resort_df, "cleaned_resort_df.csv", row.names = FALSE)

# Display the new dataframe
cleaned_resort_df

```

1. Boxplot: Ticket Prices by Continent
```{r}
# Boxplot of ticket prices by continent
ggplot(cleaned_resort_df, aes(x = Continent, y = Price, fill = Continent)) +
  geom_boxplot(outlier.shape = 21, outlier.fill = "red", outlier.size = 2) +
  labs(
    title = "Distribution of Ticket Prices by Continent",
    subtitle = "Analyzing how ticket prices vary globally",
    x = "Continent",
    y = "Ticket Price (USD)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set2")

```
The boxplot shows the distribution of ticker prices across continents, showing North America with the highest median prices but a broader range. However, the higher median price reflects the premium ski destinations. Conversely, South America and Asia have significantly lower prices, promoting these regions as an economical alternative. European resorts show a moderate median price with wide variability, emphasizing the importance of choosing carefully within the area. Ultimately, the results underscore the importance of focusing on continents with lower median prices when selecting resorts. 

2. Scatterplot: Ticket Prices vs. Highest Point
```{r}
ggplot(cleaned_resort_df, aes(x = `Highest point`, y = Price)) +
  geom_point(aes(color = Continent), alpha = 0.7, size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "blue", linetype = "dashed") +
  labs(
    title = "Ticket Prices vs. Elevation (Highest Point)",
    subtitle = "Is elevation a significant driver of ticket prices?",
    x = "Highest Point (m)",
    y = "Ticket Price (USD)",
    caption = "Data source: Cleaned Resort Dataset"
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal(base_size = 12) +
  theme(legend.title = element_text(size = 10))
```
The scatterplot visualizes a positive relationship between the highest point and ticket prices, with a trendline to confirm the pattern. The correlation implies that higher-altitude resorts, which probably have better snow conditions and more advanced terrain, charge higher prices. The difference is also available by continent, with North American and European resorts being more expensive. Therefore, cost-saving strategies would encourage one to ski outside of these regions and avoid the high-elevation resorts. 

3. Bar Chart: Average Ticket Prices by Season
```{r}
# Group seasons into broader categories
cleaned_resort_df <- cleaned_resort_df %>%
  mutate(Season_Group = case_when(
    grepl("December|November|March|April", Season) ~ "Winter",
    grepl("June|July|August", Season) ~ "Summer",
    grepl("Year-round", Season) ~ "Year-round",
    TRUE ~ "Other"
  ))


# Horizontal bar chart using grouped seasons
ggplot(cleaned_resort_df, aes(x = reorder(Season_Group, Price, FUN = mean), y = Price, fill = Season_Group)) +
  stat_summary(fun = mean, geom = "bar", color = "black", show.legend = FALSE) +
  coord_flip() + # Flip axes for horizontal bars
  labs(
    title = "Average Ticket Prices by Ski Season Group",
    x = "Season Group",
    y = "Average Ticket Price (USD)",
    caption = "Data source: Cleaned Resort Dataset"
  ) +
  theme_minimal(base_size = 12) +
  scale_fill_brewer(palette = "Set3") # Adapted color palette
```
The bar chart shows the variation in the average ticket price across different season groups. The year-round resorts have the highest average prices, emphasizing a premium for offering year-round access. Meanwhile, the summer-focussed resorts are slightly cheaper than the winter ones, probably caused by less demand between skiing seasons. The “Other” category encompasses the mixed-season resorts with the lowest average price, making them the best economical choice. The graphic helps travelers focus on specific season types to balance the cost against their skiing preferences. 


4. Heatmap: Ticket Prices by Country
```{r, fig.width=10, fig.height=15}
# Separate graphics for each continent, grouped by country
ggplot(cleaned_resort_df, aes(x = reorder(Country, Price, FUN = median), y = Price, fill = Continent)) +
  geom_boxplot(outlier.shape = 21, outlier.color = "black", outlier.size = 1) +
  coord_flip() +
  labs(
    title = "Ticket Prices by Country (Grouped by Continent)",
    x = "Country",
    y = "Ticket Price (USD)",
    caption = "Data source: Cleaned Resort Dataset"
  ) +
  theme_minimal(base_size = 12) +
  scale_fill_brewer(palette = "Set3", guide = "none") + # No legend needed since each continent is separate
  theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 10),
    strip.text = element_text(size = 12, face = "bold")
  ) +
  facet_wrap(~Continent, scales = "free_y", ncol = 1) # Facet by continent, free Y-axis scaling for clarity

```
This boxplot helps us further break down ticket prices by country to provide granular insight into the data. Higher-priced countries such as Australia, the United States, and Canada dominate the top of the list, aligning with their reputation for premium resorts. Conversely, China, Turkey, and Georgia have the lowest ticket prices, making them attractive to budget-conscious travelers. However, it is essential to consider the cost of travel rather than just the resort when making economical skiing decisions. The variability can be substantial within regions, as seen within Europe. Ultimately, it is essential to be selective of a region itself, and selectivity within that region will yield the best, most cost-effective result. 


Regression Table
```{r}
model <- lm(Price ~ `Highest point` + `Lift capacity` + `Child friendly` + Snowparks + Nightskiing, 
            data = cleaned_resort_df)

tidy_model <- tidy(model, conf.int = TRUE)

# Add a column for significance levels
tidy_model <- tidy_model %>%
  mutate(Significance = case_when(
    p.value < 0.01 ~ "***",  # 1% significance level
    p.value < 0.05 ~ "**",   # 5% significance level
    p.value < 0.1 ~ "*",     # 10% significance level
    TRUE ~ ""                # Not significant
  ))

kable(
  tidy_model,
  col.names = c("Term", "Estimate", "Std. Error", "t-value", "P-value", "Lower CI", "Upper CI", "Significance"),
  digits = 3,
  caption = "Regression Analysis of Ticket Prices: Key Predictors (Significance Levels)"
)


```
The regression results show that among all the predictors looked at, the Highest Point and the presence of Snowparks are the most significant drivers of ticket prices, with a solid positive coefficient. This highlights that resorts at higher altitudes tend to have higher prices, reflecting a premium associated with the elevation and potentially more challenging terrain. Resorts offering snowparks, which can attract freestyle or advanced skiers, can also afford the luxury of charging higher prices because of these amenities. Other variables like Lift Capacity, Child-Friendly Features, and Nightskiing do not significantly affect ticket prices. From here, the conclusion can be drawn that amenities geared toward families or casual skiing may not strongly affect price, and if, as a beginner skier, you want to avoid high prices, avoid resorts with high elevations and snowparks. 
