---
title: "Ski Resort Data Analysis - Final Paper"
author: "Arnav, Sam, Quinn, and Santiago"
date: today
format: pdf
editor: source
---


```{r, message=FALSE}
#| label: Installing-packages
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(dplyr)
library(knitr)
library(geosphere)  # For Haversine distance calculations

```

# Reading and Cleaning Data
```{r}
#| label: reading-data
resort_df <- read_csv("resorts.csv")
snow_df <- read_csv("snow.csv")
metadata <- read_csv("data_dictionary.csv")

head(resort_df)
head(snow_df)
head(metadata)

```



Dataset was found on https://www.mavenanalytics.io/data-playground?order=date_added%2Cdesc&search=ski%20resort. The csvs have been generated through webscraping from https://ski-resort-stats.com/ and the snow data come from https://nsidc.org/data/mod10cm/versions/61.

```{r}
#| label: cleaning-data
clean_non_alphanumeric <- function(data_col) {
  # Convert non-ASCII characters to ASCII
  data_col <- iconv(data_col, from = "UTF-8", to = "ASCII", sub = "")
  # Remove non-alphanumeric characters except hyphens
  data_col <- gsub("[^a-zA-Z0-9 \\-]", "", data_col)
  # Remove extra spaces
  data_col <- gsub("\\s+", " ", data_col)
  # Trim leading and trailing spaces
  trimws(data_col)
}

# Create a new dataframe with only the cleaned Resort column
cleaned_resort_df <- data.frame(CleanedResort = clean_non_alphanumeric(resort_df$Resort))

# Add the original Resort names for comparison
cleaned_resort_df$OriginalResort <- resort_df$Resort

# Write the cleaned data to a new CSV file
write.csv(cleaned_resort_df, "cleaned_resort_names.csv", row.names = FALSE)

resort_df <- resort_df |>
  mutate(clean_resort_name = clean_non_alphanumeric(Resort))
```

```{r}
#| label: merging-datasets
#plot the latitude and longitude of resorts as well as the latitude and longitude in the snow dataset with a different color
# Extract unique latitude and longitude from the snow dataset
unique_snow_df <- snow_df %>%
  distinct(Latitude, Longitude)

# Plot the latitude and longitude of resorts and snow observation locations over a world map
ggplot() +
  # Add the world map as a base layer
  borders("world", colour = "gray95", fill = "gray95") +
  coord_fixed(1.3) + # Set aspect ratio for better map projection
  # Plot snow observation locations
  geom_point(data = unique_snow_df, aes(x = Longitude, y = Latitude, color = "Snow Observation Locations"), 
             alpha = 0.7, size = 2, shape = 16) +
  # Plot resort locations
  geom_point(data = resort_df, aes(x = Longitude, y = Latitude, color = "Resort Locations"), 
             alpha = 0.7, size = 3, shape = 17) +
  # Customize colors for points
  scale_color_manual(
    values = c("Resort Locations" = "blue", "Snow Observation Locations" = "lightblue"),
    name = "Dataset"
  ) +
  # Add labels and styling
  labs(
    title = "Global Map of Resorts and Snow Observation Locations",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```
It is clear that there are many more snow observation locations across the whole world than just at the resort locations. When merging the 2 dataframes we only need to consider the snow observations for locations where there is a resort, so we will assign these observations with an ID that corresponds to the resort that is being measured. For snow observations at a longitude or latitude that don't contain a resort will have an ID value of NaN.

```{r}
# Rename ID column to Resort_ID
resort_df <- resort_df |>
  rename(Resort_ID = ID)
```


```{r}

# Function to assign resort IDs based on latitude and longitude with a specified tolerance
assign_resort_ids <- function(snow_data, resort_data, tolerance = 0.1) {
  # Initialize Resort_ID as a character column with empty strings
  snow_data <- snow_data %>%
    mutate(Resort_ID = "")
  
  for (i in seq_len(nrow(resort_data))) {
    # Get the latitude and longitude of the current resort
    resort_lat <- resort_data$Latitude[i]
    resort_lon <- resort_data$Longitude[i]
    resort_id <- resort_data$Resort_ID[i]
    
    # Filter snow data within the specified tolerance
    snow_data <- snow_data %>%
      mutate(
        Resort_ID = ifelse(
          Latitude >= (resort_lat - tolerance) & Latitude <= (resort_lat + tolerance) &
          Longitude >= (resort_lon - tolerance) & Longitude <= (resort_lon + tolerance),
          ifelse(
            Resort_ID == "",
            as.character(resort_id),  # Assign the first ID
            paste0(Resort_ID, ", ", resort_id)  # Append additional IDs
          ),
          Resort_ID
        )
      )
  }
  
  # Replace empty strings with NA for unmatched rows
  snow_data <- snow_data %>%
    mutate(Resort_ID = ifelse(Resort_ID == "", NA, Resort_ID))
  
  return(snow_data)
}

# Specify file paths for saving/loading
snow_df2_file <- "snow_df2.csv"
avg_snow_file <- "avg_snow_by_resort_and_month.csv"

# Check if `snow_df2` already exists
if (!file.exists(snow_df2_file)) {
  # Create `snow_df2` with Resort_ID assignment
  snow_df2 <- snow_df %>%
    mutate(Resort_ID = NA)
  
  # Assign Resort_IDs to snow_df
  tolerance_value <- 0.5  # Set your desired tolerance
  snow_df2 <- assign_resort_ids(snow_df2, resort_df, tolerance = tolerance_value)
  
  # Save `snow_df2` to a CSV file
  write_csv(snow_df2, snow_df2_file)
} else {
  # Load `snow_df2` from CSV
  snow_df2 <- read_csv(snow_df2_file)
}

# Create `avg_snow_by_resort_and_month` dataframe
if (!file.exists(avg_snow_file)) {
  avg_snow_by_resort_and_month <- snow_df2 %>%
    filter(!is.na(Resort_ID)) %>%
    group_by(Resort_ID, Month) %>%
    summarise(
      avg_snow = mean(Snow, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
     # Split Resort_ID into components and take the first ID for sorting
     first_resort_id = as.numeric(str_extract(Resort_ID, "^\\d+"))
    ) %>%
    arrange(first_resort_id, Resort_ID) %>% # Sort by the first ID and Resort_ID alphabetically
    select(-first_resort_id)  # Remove helper column after sorting
  
  # Save `avg_snow_by_resort_and_month` to a CSV file
  write_csv(avg_snow_by_resort_and_month, avg_snow_file)
} else {
  avg_snow_by_resort_and_month <- read_csv(avg_snow_file)
}

# Ensure Resort_ID is numeric in avg_snow_by_resort_and_month
all_resort_ids <- seq(1, 499)
present_resort_ids <- unique(as.numeric(unlist(strsplit(avg_snow_by_resort_and_month$Resort_ID, ",\\s*"))))

# Find missing Resort_IDs
missing_resort_ids <- setdiff(all_resort_ids, present_resort_ids)

# Print the missing IDs
print(missing_resort_ids) # This is the resort that has no snow data

```


```{r}
# Count unique latitude and longitude values
unique_latitudes <- snow_df %>%
  summarise(unique_lat = n_distinct(Latitude)) %>%
  pull(unique_lat)

unique_longitudes <- snow_df %>%
  summarise(unique_lon = n_distinct(Longitude)) %>%
  pull(unique_lon)

cat("Number of unique latitudes:", unique_latitudes, "\n")
cat("Number of unique longitudes:", unique_longitudes, "\n")

# Bin the latitude and longitude, group by Month, and calculate average Snow
binned_snow_df <- snow_df %>%
  mutate(
    lat_bin = floor(Latitude/5)*5,          # Bin latitude to nearest lower degree
    lon_bin = floor(Longitude / 8) * 8 # Bin longitude to nearest lower 5 degrees
  ) %>%
  group_by(Month, lat_bin, lon_bin) %>%
  summarise(
    avg_snow = mean(Snow, na.rm = TRUE), # Average Snow for the bin
    .groups = "drop"                     # Ungroup after summarise
  )

# View the resulting data frame
print(binned_snow_df)

```

# Exploratory Data Analysis

```{r}

```


```{r}
generate_avg_bar_chart <- function(data, group_col, avg_col, title, x_label, y_label, fill_color = "steelblue") {
  ggplot(data, aes_string(x = group_col, y = avg_col)) +
    stat_summary(fun = "mean", geom = "bar", fill = fill_color, color = "black") +
    labs(title = title, x = x_label, y = y_label) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = ifelse(group_col == "Country", 45, 0), hjust = 1))
}


avg_price_continent <- generate_avg_bar_chart(
  data = resort_df,
  group_col = "Continent",
  avg_col = "Price",
  title = "Average Price per Continent",
  x_label = "Continent",
  y_label = "Average Price",
  fill_color = "steelblue"
)

print(avg_price_continent)


avg_price_country <- generate_avg_bar_chart(
  data = resort_df,
  group_col = "Country",
  avg_col = "Price",
  title = "Average Price per Country",
  x_label = "Country",
  y_label = "Average Price",
  fill_color = "slategrey"
)

print(avg_price_country)

```


```{r}
# Function to create a bar chart for counts
generate_count_bar_chart <- function(data, group_col, title, x_label, y_label, fill_color = "steelblue") {
  ggplot(data, aes_string(x = group_col)) +
    geom_bar(fill = fill_color, color = "black") +
    labs(title = title, x = x_label, y = y_label) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = ifelse(group_col == "Country", 45, 0), hjust = 1))
}

# Generate bar chart for count of resorts per continent
count_resorts_continent <- generate_count_bar_chart(
  data = resort_df,
  group_col = "Continent",
  title = "Count of Resorts per Continent",
  x_label = "Continent",
  y_label = "Count of Resorts",
  fill_color = "steelblue"
)

# Display the first chart
print(count_resorts_continent)

# Generate bar chart for count of resorts per country
count_resorts_country <- generate_count_bar_chart(
  data = resort_df,
  group_col = "Country",
  title = "Count of Resorts per Country",
  x_label = "Country",
  y_label = "Count of Resorts",
  fill_color = "coral"
)

# Display the second chart
print(count_resorts_country)

```


```{r}
library(patchwork)
library(ggplot2)
library(dplyr)

# Calculate proportions for each feature
child_friendly_data <- resort_df %>%
  summarise(Yes = sum(`Child friendly` == "Yes", na.rm = TRUE),
            No = sum(`Child friendly` == "No", na.rm = TRUE)) %>%
  gather(key = "Response", value = "Count") %>%
  mutate(Percentage = Count / sum(Count) * 100)

snowparks_data <- resort_df %>%
  summarise(Yes = sum(Snowparks == "Yes", na.rm = TRUE),
            No = sum(Snowparks == "No", na.rm = TRUE)) %>%
  gather(key = "Response", value = "Count") %>%
  mutate(Percentage = Count / sum(Count) * 100)

summer_data <- resort_df %>%
  summarise(Yes = sum(`Summer skiing` == "Yes", na.rm = TRUE),
            No = sum(`Summer skiing` == "No", na.rm = TRUE)) %>%
  gather(key = "Response", value = "Count") %>%
  mutate(Percentage = Count / sum(Count) * 100)

nightskiing_data <- resort_df %>%
  summarise(Yes = sum(Nightskiing == "Yes", na.rm = TRUE),
            No = sum(Nightskiing == "No", na.rm = TRUE)) %>%
  gather(key = "Response", value = "Count") %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Helper function to create pie charts
create_pie_chart <- function(data, title) {
  ggplot(data, aes(x = "", y = Percentage, fill = Response)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar(theta = "y") +
    theme_void() +
    labs(title = title, fill = "Response") +
    theme(plot.title = element_text(hjust = 0.5))
}

# Generate individual pie charts
child_friendly_plot <- create_pie_chart(child_friendly_data, "Child Friendly")
snowparks_plot <- create_pie_chart(snowparks_data, "Availability of Snowparks")
summer_plot <- create_pie_chart(summer_data, "Availability of Summer Skiing")
nightskiing_plot <- create_pie_chart(nightskiing_data, "Availability of Nightskiing")

# Combine the pie charts into one plot
combined_plot <- (child_friendly_plot | snowparks_plot) / (summer_plot | nightskiing_plot)

# Display the combined plot
print(combined_plot)
```
```{r}
resort_df$slope_level_ratios <- paste(
  round(resort_df$`Beginner slopes` / resort_df$`Total slopes`, 1),
  round(resort_df$`Intermediate slopes` / resort_df$`Total slopes`, 1),
  round(resort_df$`Difficult slopes` / resort_df$`Total slopes`, 1),
  sep = "-"
)

# View the updated dataset
head(resort_df)


#calculating the average ratios in the top 10 countries with the most resorts
#displaying the average ratio breakup in those countries


```











```{r}
# Define the mapping from month names to numeric values
month_to_number <- c(
  "January" = 1,
  "February" = 2,
  "March" = 3,
  "April" = 4,
  "May" = 5,
  "June" = 6,
  "July" = 7,
  "August" = 8,
  "September" = 9,
  "October" = 10,
  "November" = 11,
  "December" = 12
)

# Function to convert a single season range (e.g., "December - April") into numeric months
convert_range_to_months <- function(range) {
  months <- strsplit(range, " - ")[[1]]
  start_month <- months[1]
  end_month <- months[2]
  
  start_num <- month_to_number[[start_month]]
  end_num <- month_to_number[[end_month]]
  
  if (start_num <= end_num) {
    return(start_num:end_num)
  } else {
    return(c(start_num:12, 1:end_num))
  }
}

# Function to handle multiple ranges in a season string
convert_season_to_months_multiple <- function(season) {
  # Split the season string by commas
  ranges <- strsplit(season, ", ")[[1]]
  
  # Convert each range to numeric months and combine the results
  all_months <- unlist(lapply(ranges, convert_range_to_months))
  
  # Remove duplicates and return sorted months
  return(sort(unique(all_months)))
}

# Apply the function to the "Season" column
resort_df$Season_Months <- lapply(resort_df$Season, convert_season_to_months_multiple)

# Print the dataset with the new column
print(resort_df)

```
