---
title: "Ski Resort Data Analysis - Final Paper"
author: "Arnav, Sam, Quinn, and Santiago"
date: today
format: pdf
editor: source
---


```{r, message=FALSE}
#| label: Installing-packages
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(dplyr)
library(knitr)
library(geosphere)  # For Haversine distance calculations

```

# Reading and Cleaning Data
```{r}
#| label: reading-data
resort_df <- read_csv("resorts.csv")
snow_df <- read_csv("snow.csv")
metadata <- read_csv("data_dictionary.csv")

head(resort_df)
head(snow_df)
head(metadata)

```



Dataset was found on https://www.mavenanalytics.io/data-playground?order=date_added%2Cdesc&search=ski%20resort. The csvs have been generated through webscraping from https://ski-resort-stats.com/ and the snow data come from https://nsidc.org/data/mod10cm/versions/61.

```{r}
#| label: cleaning-data
clean_non_alphanumeric <- function(data_col) {
  # Convert non-ASCII characters to ASCII
  data_col <- iconv(data_col, from = "UTF-8", to = "ASCII", sub = "")
  # Remove non-alphanumeric characters except hyphens
  data_col <- gsub("[^a-zA-Z0-9 \\-]", "", data_col)
  # Remove extra spaces
  data_col <- gsub("\\s+", " ", data_col)
  # Trim leading and trailing spaces
  trimws(data_col)
}

# Create a new dataframe with only the cleaned Resort column
cleaned_resort_df <- data.frame(CleanedResort = clean_non_alphanumeric(resort_df$Resort))

# Add the original Resort names for comparison
cleaned_resort_df$OriginalResort <- resort_df$Resort

# Write the cleaned data to a new CSV file
write.csv(cleaned_resort_df, "cleaned_resort_names.csv", row.names = FALSE)

cleaned_resort_df
```

```{r}
#| label: merging-datasets
#plot the latitude and longitude of resorts as well as the latitude and longitude in the snow dataset with a different color
# Extract unique latitude and longitude from the snow dataset
unique_snow_df <- snow_df %>%
  distinct(Latitude, Longitude)

# Plot the latitude and longitude of resorts and snow observation locations over a world map
ggplot() +
  # Add the world map as a base layer
  borders("world", colour = "gray95", fill = "gray95") +
  coord_fixed(1.3) + # Set aspect ratio for better map projection
  # Plot snow observation locations
  geom_point(data = unique_snow_df, aes(x = Longitude, y = Latitude, color = "Snow Observation Locations"), 
             alpha = 0.7, size = 2, shape = 16) +
  # Plot resort locations
  geom_point(data = resort_df, aes(x = Longitude, y = Latitude, color = "Resort Locations"), 
             alpha = 0.7, size = 3, shape = 17) +
  # Customize colors for points
  scale_color_manual(
    values = c("Resort Locations" = "blue", "Snow Observation Locations" = "lightblue"),
    name = "Dataset"
  ) +
  # Add labels and styling
  labs(
    title = "Global Map of Resorts and Snow Observation Locations",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```
It is clear that there are many more snow observation locations across the whole world than just at the resort locations. When merging the 2 dataframes we only need to consider the snow observations for locations where there is a resort, so we will assign these observations with an ID that corresponds to the resort that is being measured. For snow observations at a longitude or latitude that don't contain a resort will have an ID value of NaN.

```{r}
# Rename ID column to Resort_ID
resort_df <- resort_df %>%
  rename(Resort_ID = ID)

# Count unique latitude and longitude values
unique_latitudes <- snow_df %>%
  summarise(unique_lat = n_distinct(Latitude)) %>%
  pull(unique_lat)

unique_longitudes <- snow_df %>%
  summarise(unique_lon = n_distinct(Longitude)) %>%
  pull(unique_lon)

cat("Number of unique latitudes:", unique_latitudes, "\n")
cat("Number of unique longitudes:", unique_longitudes, "\n")

```

# Exploratory Data Analysis

```{r}

```


```{r}
generate_avg_bar_chart <- function(data, group_col, avg_col, title, x_label, y_label, fill_color = "steelblue") {
  ggplot(data, aes_string(x = group_col, y = avg_col)) +
    stat_summary(fun = "mean", geom = "bar", fill = fill_color, color = "black") +
    labs(title = title, x = x_label, y = y_label) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = ifelse(group_col == "Country", 45, 0), hjust = 1))
}


avg_price_continent <- generate_avg_bar_chart(
  data = resort_df,
  group_col = "Continent",
  avg_col = "Price",
  title = "Average Price per Continent",
  x_label = "Continent",
  y_label = "Average Price",
  fill_color = "steelblue"
)

print(avg_price_continent)


avg_price_country <- generate_avg_bar_chart(
  data = resort_df,
  group_col = "Country",
  avg_col = "Price",
  title = "Average Price per Country",
  x_label = "Country",
  y_label = "Average Price",
  fill_color = "slategrey"
)

print(avg_price_country)

```


```{r}
# Function to create a bar chart for counts
generate_count_bar_chart <- function(data, group_col, title, x_label, y_label, fill_color = "steelblue") {
  ggplot(data, aes_string(x = group_col)) +
    geom_bar(fill = fill_color, color = "black") +
    labs(title = title, x = x_label, y = y_label) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = ifelse(group_col == "Country", 45, 0), hjust = 1))
}

# Generate bar chart for count of resorts per continent
count_resorts_continent <- generate_count_bar_chart(
  data = resort_df,
  group_col = "Continent",
  title = "Count of Resorts per Continent",
  x_label = "Continent",
  y_label = "Count of Resorts",
  fill_color = "steelblue"
)

# Display the first chart
print(count_resorts_continent)

# Generate bar chart for count of resorts per country
count_resorts_country <- generate_count_bar_chart(
  data = resort_df,
  group_col = "Country",
  title = "Count of Resorts per Country",
  x_label = "Country",
  y_label = "Count of Resorts",
  fill_color = "coral"
)

# Display the second chart
print(count_resorts_country)

```


```{r}
library(patchwork)

# Helper function to create pie charts
create_pie_chart <- function(data, title) {
  df <- as.data.frame(data)
  colnames(df) <- c("Response", "Percentage")
  ggplot(df, aes(x = "", y = Percentage, fill = Response)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar(theta = "y") +
    theme_void() +
    labs(title = title, fill = "Response") +
    theme(plot.title = element_text(hjust = 0.5))
}

# Generate individual pie charts
child_friendly_plot <- create_pie_chart(child_friendly_data, "Child Friendly")
snowparks_plot <- create_pie_chart(snowparks_data, "Availability of Snowparks")
summer_plot <- create_pie_chart(summer_data, "Availability of Summer Skiing")
nightskiing_plot <- create_pie_chart(nightskiing_data, "Availability of Nightskiing")

# Combine the pie charts into one plot
combined_plot <- (child_friendly_plot | snowparks_plot) / (summer_plot | nightskiing_plot)

# Display the combined plot
print(combined_plot)
```
```{r}
resort_df$slope_level_ratios <- paste(
  round(resort_df$`Beginner slopes` / resort_df$`Total slopes`, 1),
  round(resort_df$`Intermediate slopes` / resort_df$`Total slopes`, 1),
  round(resort_df$`Difficult slopes` / resort_df$`Total slopes`, 1),
  sep = "-"
)

# View the updated dataset
head(resort_df)


#calculating the average ratios in the top 10 countries with the most resorts
#displaying the average ratio breakup in those countries


```

