---
title: "Ski Resort Data Analysis - Final Paper"
author: "Arnav, Sam, Quinn, and Santiago"
date: today
format: pdf
editor: source
---


```{r, message=FALSE}
#| label: Installing-packages
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(dplyr)
library(knitr)
library(geosphere)  # For Haversine distance calculations

```

# Reading and Cleaning Data
```{r}
#| label: reading-data
resort_df <- read_csv("resorts.csv")
snow_df <- read_csv("snow.csv")
metadata <- read_csv("data_dictionary.csv")

head(resort_df)
head(snow_df)
head(metadata)

view(metadata)

```

Dataset was found on https://www.mavenanalytics.io/data-playground?order=date_added%2Cdesc&search=ski%20resort. The csvs have been generated through webscraping from https://ski-resort-stats.com/ and the snow data come from https://nsidc.org/data/mod10cm/versions/61.

```{r}
#| label: cleaning-data

```

```{r}
#| label: merging-datasets
#plot the latitude and longitude of resorts as well as the latitude and longitude in the snow dataset with a different color

# Extract unique latitude and longitude from the snow dataset
unique_snow_df <- snow_df %>%
  distinct(Latitude, Longitude)

# Plot the latitudes and longitudes with separate layers and a legend
ggplot() +
  geom_point(data = unique_snow_df, aes(x = Longitude, y = Latitude, color = "Snow Observation Locations"), 
             alpha = 0.7, size = 2, shape = 16) +
  geom_point(data = resort_df, aes(x = Longitude, y = Latitude, color = "Resort Locations"), 
             alpha = 0.7, size = 3, shape = 17) +
  scale_color_manual(
    values = c("Resort Locations" = "blue", "Snow Observation Locations" = "red"),
    name = "Dataset"
  ) +
  labs(
    title = "Resorts and Snow Data: Latitude and Longitude",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```
It is clear that there are many more snow observation locations across the whole world than just at the resort locations. When merging the 2 dataframes we only need to consider the snow observations for locations where there is a resort, so we will assign these observations with an ID that corresponds to the resort that is being measured. For snow observations at a longitude or latitude that don't contain a resort will have an ID value of NaN.

```{r}
# Rename ID column to Resort_ID
resort_df <- resort_df %>%
  rename(Resort_ID = ID)

# Function to assign resort IDs to snow observations
assign_resort_id <- function(snow_data, resort_data, tolerance_km = 10) {
  # Initialize a column for the closest Resort_ID
  snow_data$Resort_ID <- NA
  
  # Loop through each snow observation
  for (i in seq_len(nrow(snow_data))) {
    # Extract latitude and longitude of the snow observation
    snow_lat <- snow_data$Latitude[i]
    snow_lon <- snow_data$Longitude[i]
    
    # Calculate distances to all resorts
    distances <- distHaversine(
      cbind(resort_data$Longitude, resort_data$Latitude),
      c(snow_lon, snow_lat)
    ) / 1000  # Convert meters to kilometers
    
    # Check if any resort is within the tolerance
    if (any(distances <= tolerance_km)) {
      # Assign the ID of the closest resort
      snow_data$Resort_ID[i] <- resort_data$Resort_ID[which.min(distances)]
    } else {
      # No resort within tolerance, assign NA
      snow_data$Resort_ID[i] <- NA
    }
  }
  
  return(snow_data)
}

# Match the snow observations to the closest resort within a 10 km tolerance
matched_snow_df <- assign_resort_id(snow_df, resort_df, tolerance_km = 10)

# View the resulting dataset
print(head(matched_snow_df))

# Summarize matched and unmatched observations
summary_table <- matched_snow_df %>%
  summarise(
    Total_Observations = n(),
    Matched_Observations = sum(!is.na(Resort_ID)),
    Unmatched_Observations = sum(is.na(Resort_ID))
  )

# Display the summary
print(summary_table)

# Visualize the matched data
ggplot() +
  geom_point(data = resort_df, aes(x = Longitude, y = Latitude), color = "blue", size = 3, alpha = 0.7, shape = 16) +
  geom_point(data = matched_snow_df, aes(x = Longitude, y = Latitude, color = !is.na(Resort_ID)), 
             alpha = 0.7, size = 2, shape = 17) +
  scale_color_manual(
    values = c("TRUE" = "green", "FALSE" = "red"),
    labels = c("Matched", "Unmatched"),
    name = "Snow Data Match"
  ) +
  labs(
    title = "Matching Snow Observations to Resorts",
    x = "Longitude",
    y = "Latitude",
    caption = "Green: Matched | Red: Unmatched | Blue: Resorts"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# Exploratory Data Analysis

```{r}

```

