---
title: "GYM Data Analysis"
author: "Arnav, Sam, Quinn, and Santiago"
format: pdf
editor: visual
data: today
title: 11/13/24
---

# Question 1

For lab credit, submit at least one table or figure from (some of?) the data that you plan to use for the final\
project. Describe the data in the figure or the values in the table.\

Choosing your research question and data is a very important step. Try to choose something that you care\
about and that you can use to practice and further develop your data science skills.\

```{r, message=FALSE}
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(dplyr)
```

With this data set we are considering taking the approach of "being" a personal trainer who is analyzing this data set in order to best determine what types of lessons to offer for what specific person.

```{r}

<<<<<<< Updated upstream
gym_members_data <- read_csv("gym_members_exercise_tracking.csv")

mega_gym_data <- read_csv("megaGymDataset.csv")

head(gym_members_data)
head(mega_gym_data)
```

```{r}

type_mega_counts <- mega_gym_data %>%
  count(Type) %>%
  arrange(desc(n))

# View the results
type_mega_counts


type_gym_counts <- gym_members_data %>%
  count(Workout_Type) %>%
  arrange(desc(n))

# View the results
type_gym_counts
```

in these two figures above we had each data set be counted by the type of workout. What we can see is that for Mega Counts strength workouts are a lot more common with 2,545 followed by stretching with only 147. This story continues for the Gym Counts as strength is the most common one but not by nearly as large as a margin as we see that the count is 258 and is followed by Cardio with 255. The two graphics below help display this better.
