---
title: "GYM Data Analysis"
author: "Arnav, Sam, Quinn, and Santiago"
date: today
format: pdf
editor: source
---

# Question 1

For lab credit, submit at least one table or figure from (some of?) the data that you plan to use for the final
project. Describe the data in the figure or the values in the table.

Choosing your research question and data is a very important step. Try to choose something that you care
about and that you can use to practice and further develop your data science skills.

```{r, message=FALSE}
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(dplyr)
```

With this data set we are considering taking the approach of "being" a personal trainer who is analyzing this data set in order to best determine what types of lessons to offer for what specific person.

```{r}

gym_members_data <- read_csv("gym_members_exercise_tracking.csv")

mega_gym_data <- read_csv("megaGymDataset.csv")

head(gym_members_data)
head(mega_gym_data)
```

```{r}

type_mega_counts <- mega_gym_data %>%
  count(Type) %>%
  arrange(desc(n))

# View the results
type_mega_counts


type_gym_counts <- gym_members_data %>%
  count(Workout_Type) %>%
  arrange(desc(n))

# View the results
type_gym_counts
```

```{r}


# Boxplot of Calories Burned by the specified exercise types
ggplot(gym_members_data, aes(x = Workout_Type, y = Calories_Burned)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Calories Burned by Selected Exercise Types", x = "Exercise Type", y = "Calories Burned") +
  theme_minimal()
```


in these two figures above we had each data set be counted by the type of workout. What we can see is that for Mega Counts strength workouts are a lot more common with 2,545 followed by stretching with only 147. This story continues for the Gym Counts as strength is the most common one but not by nearly as large as a margin as we see that the count is 258 and is followed by Cardio with 255. The two graphics below help display this better.

```{r}

ggplot(type_mega_counts, aes(x = reorder(Type, -n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Workout Type Counts in Mega Gym Data",
       x = "Workout Type",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Plot for type_gym_counts
ggplot(type_gym_counts, aes(x = reorder(Workout_Type, -n), y = n)) +
  geom_bar(stat = "identity", fill = "darkorange") +
  labs(title = "Workout Type Counts in Gym Members Data",
       x = "Workout Type",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Research Question

How does the type of exercise and experience level influence calorie burn and heart rate metrics across different age groups and genders, and is there a correlation with the exercise ratings provided by gym members?

